Spatial and temporal decay of SARS-CoV-2 environmental contamination of the healthcare environment

### 2021-09-28 manuscript.Rmd
Manuscript Revision

## Authors  
Matthew J. Ziegler, MD, MS (1,2,3); Selamawit Bekele (2,3); Sean Loughrey, MS (2,3); Emily Reesey, MS (2,2); Elizabeth Huang (2,3); Pam Tolomeo, MPH (2,3); Ebbing Lautenbach, MD, MPH, MS (1,2,3); Brendan J. Kelly, MD, MS (1,2,3) for the CDC Prevention Epicenters Program

@@

## Authors' Contributions  
MJZ - study design, data collection and analysis, manuscript;  
SB - subject enrollment and specimen collection;  
SL - specimen processing and analysis;  
ER - specimen processing and analysis;  
EH - specimen collection and processing;   
PT - subject enrollment and specimen collection, manuscript;  
MZD - study design, manuscript;  
EL - study design, manuscript;  
BJK - study design, data collection and analysis, manuscript;

@@

* __Background__: SARS-CoV-2 is known to contaminate surfaces within healthcare settings, but the spatial and temporal extent of environmental contamination has not been precisely defined. We sought to define SARS-CoV-2 contamination of different surface types and how contamination changes over time.

* __Methods__: We performed longitudinal sampling of surfaces within COVID-19 patient rooms while measuring the distance from each surface to the patient bed. We performed quantitative RT-PCR for the detection of SARS-CoV-2 RNA and modeled the role of distance, time, and severity of illness on the probability of detecting SARS-CoV-2 using a mixed-effects binomial model.

* __Results__: The probability of detecting SARS-CoV-2 RNA did not vary with distance. However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors odds ratio (OR) `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)). Patients requiring high-flow oxygen, positive airway pressure, or mechanical ventilation respiratory were associated with increased surface contamination (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)). The probability of surface elevated contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.

* __Conclusions__: Distance from patient's bed did not predict SARS-CoV-2 RNA deposition in patient rooms, but surface type, severity of illness, and time predicted surface deposition.

@@

Despite infection control measures, transmission of SARS-CoV-2 infection has been observed in the healthcare and hospital settings. In a recent study of SARS-CoV-2-infected and -uninfected patients inadvertently roomed together, 39% of exposed uninfected patients developed COVID-19. Risk of SARS-CoV-2 transmission was associated with source patients who had low cycle threshold (Ct), i.e., high viral load, in their diagnostic testing [@Karan2021-cm].

SARS-CoV-2 transmission can be mediated by droplets, aerosols, or fomites. Prior studies have demonstrated that SARS-CoV-2 extensively contaminates surfaces in the healthcare environment [@Morawska2021-mv]. However, the spatial extent of contamination and how contamination changes during the course of hospitalization remain less well understood [@Mody2021-ag].

We have previously applied a systematic sampling and spatial modeling strategy to define how multidrug-resistant organisms cluster in the hospital environment [@Kelly2021-vm]. We now extend the methods we have developed to understand how SARS-CoV-2 contamination is distributed spatially, and how it changes over time.

@@

* __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA (IRB protocol #843273). Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability. All rooms underwent daily cleaning, according to hospital protocol. Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch objects, limiting use to a single surface. Floors were cleaned using a microfiber mop and sodium dichloroisocyanurate, a chlorine-based disinfectant. Patient rooms were eligible for inclusion in the study if the occupant had a positive test for SARS-CoV-2 within the prior 7 days and was admitted to a new patient room with COVID-19 isolation precautions.

@@

* __Specimen Collection and SARS-CoV-2 RT-PCR__: COVID-19 patient rooms were swabbed longitudinally until day 28 or patient discharge from the hospital, with sampling targeted to hospital days 1, 2, 7, 14, 21, and 28. Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room. Multiple surfaces were swabbed within each patient room using a flocked nylon swab (Copan) within a sterile 20 cm^2^ template to ensure consistent sampling. High-touch surfaces (**Table 1**) were selected using structured observations prior to study start. Swabs were immediately placed into buffered AVL and frozen prior to further processing. Specimens then underwent RNA extraction (<span style="background-color:cyan">which kit?</span>) and quantitative reverse transcription polymerase chain reaction (RT-qPCR) (Thermo Quantstudio) with primers and probe targeting the SARS-CoV-2 N1 region (Centers for Disease Control and Prevention (CDC) RT-qPCR probe assay, IDT) in triplicate.

* __Distance Data Collection__: Distance from the environmental sampling site to the patient's head of bed was recorded using a laser measuring device (Bosch GLM 20). Templates were affixed to each surface and the linear distance from the patient's head of bed to the sample site was measured.

* __Definition of Exposures and Outcomes__: The primary outcome of interest, detection of SARS-CoV-2 by RT-qPCR assay, was determined by detection of any quantifiable RNA on <span style="background-color:cyan">any of three technical replicates</span>. The primary exposures of interest were (1) distance from the head of patient's bed, measured as above; (2) time from COVID-19 diagnosis and from the start of the local COVID-19 case wave; and (3) COVID-19 clinical severity. We categorized COVID-19 severity by the highest level of oxygen support required during the subjects' admission, wither no oxygen support required (mild disease), only nasal or facemask oxygen support required (moderate disease), significant oxygen support (i.e., high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation) required (severe disease).

@@

* __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces. There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of multidrug-resistant bacterial organisms [@Kelly2021-vm], but the suface type was significantly associated with the probability of detection. Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to elevated surfaces. High-touch elevated surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces. **Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.

@@

We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms. We found that distance from the patient's bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surace (elevated or floor, high-touch or low-touch) did predict different levels of contamination. We also found a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis, consistent with previous reports [@Chia2020]. While the overall probability of SARS-CoV-2 RNA detection decreased as time from diagnosis increased, the probability of SARS-CoV-2 RNA detection on floors increased with time from the onset of the local COVID-19 case wave. The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or the mops of EVS teams increases with the COVID-19 patient census. This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas [@Guo2020]. Additionally, it is possible that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.

We found that increased clinical severity of COVID-19, measured by oxygen requirement, was associated with increased contamination of the healthcare environment. Specifically, those subjects who required high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation, were found to have increased odds of detecting surface SARS-CoV-2 RNA, compared to those with no oxygen requirement. This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory procedures. In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021], potentially supporting the contribution of respiratory care in surface contamination. While we can only speculate on the mechanism for greater detection in these patients, our findings support CDC guidance for prolonged isolation in critically ill patients [@CDC2021-gc].

Our study design and hypotheses were informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@Kelly2021-vm]. Contrary to our findings with multidrug-resistant bacteria, SARS-CoV-2 detection did not vary significantly within any surface group as distance from the patient increased. The difference between the findings from our two studies is most likely related to two factors: (1) the relatively small size of patient rooms and (2) the physical spread of SARS-CoV-2 RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient's room. This finding matches previous description of high probability of floor contamination relative to other surfaces [@Chia2020]. In contrast to floors, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.

Our study has several limitations. First, we used RT-qPCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces. Secondly, we did not measure differences in cleaning methods across different study units. However, cleaning practices did not undergo systematic changes during the study time period and sampling was performed routinely prior to daily cleaning. Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.

In conclusion, SARS-CoV-2 RNA surface deposition did not vary with distance within patient rooms. However, surface type, severity of illness, and time predicted detection of surface contamination. Future studies should focus on how the observed surface contamination contributes to risk of infection among patients and HCP, and on how cleaning strategies can target surfaces found to be at high risk of contamination.

>>>>

## Authors  
Matthew J. Ziegler, MD, MS (1,2,3); Elizabeth Huang (2,3); Selamawit Bekele (2,3); Emily Reesey, MS (2,2); Pam Tolomeo, MPH (2,2), Sean Loughrey, MS (2,3); Michael Z. David, MD, PhD (1,2,3); Ebbing Lautenbach, MD, MPH, MS (1,2,3); Brendan J. Kelly, MD, MS (1,2,3) for the CDC Prevention Epicenters Program

@@

## Authors' Contributions  
MJZ - study design, data collection and analysis, manuscript;
EH - specimen collection and processing;  
SB - subject enrollment and specimen collection;  
SL - specimen processing and analysis;  
ER - specimen processing and analysis;  
PT - subject enrollment and specimen collection, manuscript;  
MZD - study design, manuscript;  
EL - study design, manuscript;  
BJK - study design, data collection and analysis, manuscript;

@@

* __Background__: The spatial and temporal extent of SARS-CoV-2 environmental contamination has not been precisely defined. We sought to elucidate contamination of different surface types and how contamination changes over time.

* __Methods__: We sampled surfaces longitudinally within COVID-19 patient rooms, performed quantitative RT-PCR for the detection of SARS-CoV-2 RNA, and modeled distance, time, and severity of illness on the probability of detecting SARS-CoV-2 using a mixed-effects binomial model.

* __Results__: The probability of detecting SARS-CoV-2 RNA in a patietn room did not vary with distance. However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors odds ratio (OR) `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)). Patients requiring high-flow oxygen, positive airway pressure, or mechanical ventilation were associated with increased surface contamination (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)). The probability of surface elevated contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.

* __Conclusions__: Distance from patient’s bed did not predict SARS-CoV-2 RNA deposition in patient rooms, but surface type, severity of illness, and time from local pandemic wave predicted surface deposition.

@@

Despite infection control measures, transmission of SARS-CoV-2 infection has been observed in the hospital and other healthcare settings. In a recent study of SARS-CoV-2-infected and -uninfected patients inadvertently roomed together, 39% of exposed, previoulsy uninfected patients developed COVID-19. Risk of SARS-CoV-2 transmission was associated with source patients who had low cycle threshold (Ct), i.e., high viral load, in their diagnostic testing [@Karan2021-cm].

SARS-CoV-2 transmission can be mediated by droplets, aerosols, or fomites. Prior studies have demonstrated that SARS-CoV-2 extensively contaminates surfaces in the healthcare environment [@Morawska2021-mv; @Ong2021_1; @Ryu2020; @Moore2021; @Coil2021; @Redmond2021; @Ong2021_2]. However, the spatial extent of contamination and how contamination changes during the course of hospitalization remain less well understood [@Mody2021-ag].

We have previously applied a systematic sampling and spatial modeling strategy to define how multidrug-resistant organisms cluster in the hospital environment [@Kelly2021]. We now extend the methods we have developed to understand how SARS-CoV-2 contamination is distributed spatially, and how it changes over time.

@@

* __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA (IRB protocol #843273). Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability. All rooms underwent daily cleaning, according to hospital protocol. Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch surfaces, limiting use to a single surface. Floors were cleaned using a microfiber mop and sodium dichloroisocyanurate, a chlorine-based disinfectant. Patient rooms were eligible for inclusion in the study if the occupant had a positive test for SARS-CoV-2 within the prior 7 days and was admitted to a new patient room with COVID-19 isolation precautions.

@@

* __Specimen Collection and SARS-CoV-2 RT-PCR__: COVID-19 patient rooms were swabbed longitudinally until day 28 or patient discharge from the hospital, with sampling targeted to hospital days 1, 2, 7, 14, 21, and 28. Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room. Multiple surfaces were swabbed within each patient room using a flocked nylon swab (Copan) within a sterile 20 cm2 template to ensure consistent sampling. High-touch surfaces (Table 1) were selected using structured observations prior to study start. Swabs were immediately placed into buffered AVL and frozen prior to further processing. Specimens then underwent RNA extraction (QIAmp Viral RNA Mini Kit) and quantitative reverse transcription polymerase chain reaction (RT-qPCR) (Thermo Quantstudio) with primers and probe targeting the SARS-CoV-2 N1 region (Centers for Disease Control and Prevention (CDC) RT-qPCR probe assay, IDT) in triplicate.

* __Distance Data Collection__: Distance from the environmental sampling site to the patient’s head of bed was recorded using a laser measuring device (Bosch GLM 20). Templates were affixed to each surface and the linear distance from the patient’s head of bed to the sample site was measured.

* __Definition of Exposures and Outcomes__: The primary outcome of interest, detection of SARS-CoV-2 by RT-qPCR assay, was determined by detection of any quantifiable RNA on any of three technical replicates. Replicates were reviewed for inconsistencies and PCR was repeated in the event of disparate results. The primary exposures of interest were (1) distance from the head of patient’s bed, measured as above; (2) time from COVID-19 diagnosis and from the start of the local COVID-19 case wave; and (3) COVID-19 clinical severity. We categorized COVID-19 severity by the highest level of oxygen support required during the subjects’ admission: 1) no oxygen support required (mild disease); 2) only nasal or facemask oxygen support required (moderate disease); and 3) high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation (severe disease).

@@

* __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces. There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of multidrug-resistant bacterial organisms [@Kelly2021], but the suface type was significantly associated with the probability of detection. Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to elevated surfaces. High-touch elevated surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces. **Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.

@@

We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms. We found that distance from the patient’s bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surface (elevated or floor, high-touch or low-touch) did predict different levels of contamination with highest probability among floors followed be elevated high-touch surfaces. We also found a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis, consistent with previous reports [@Chia2020]. While the overall probability of SARS-CoV-2 RNA detection decreased as time from diagnosis increased, the probability of SARS-CoV-2 RNA detection on floors increased with time from the onset of the local COVID-19 case wave. The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or the mops of EVS teams increases with the COVID-19 patient census. This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas[@Guo2020]. Additionally, it is possible that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.

We found that increased clinical severity of COVID-19, measured by oxygen requirement, was associated with increased contamination of the healthcare environment. Specifically, those subjects who required high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation, were found to have increased odds of detecting surface SARS-CoV-2 RNA, compared to those with no oxygen requirement. This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory procedures. In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021], potentially supporting the contribution of respiratory care in surface contamination. While we can only speculate on the mechanism for greater detection in these patients, our findings support CDC guidance for prolonged isolation in critically ill patients[@CDC2021].

Our study design and hypotheses were informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room[@Kelly2021]. Contrary to our findings with multidrug-resistant bacteria, SARS-CoV-2 detection did not vary significantly within any surface group as distance from the patient increased. The difference between the findings from our two studies is most likely related to two factors: (1) the relatively small size of patient rooms and (2) the physical spread of SARS-CoV-2 RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient’s room. This finding matches previous description of high probability of floor contamination relative to other surfaces[@Chia2020]. In contrast to floors, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.

Our study has several limitations. First, we used RT-qPCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces. Secondly, we did not measure differences in cleaning methods across different study units. However, cleaning practices did not undergo systematic changes during the study period and sampling was performed routinely prior to daily cleaning. Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.

In conclusion, SARS-CoV-2 RNA surface deposition did not vary with distance within patient rooms. However, surface type, severity of illness, and time from beginning of the local COVID-19 wave predicted detection of surface contamination. Future studies should focus on how the observed surface contamination contributes to risk of infection among patients and HCP, and on how cleaning strategies can target surfaces found to be at high risk of contamination.
